---
title: "Linear Modeling"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Inference about a mean

A new reading program is being evaluated at an elementary school. A random sample of 20 students were tested to determine reading speed. Speed was measured in minutes. [@ott2004]

```{r}
speed <- c(5, 7, 15, 12, 8, 7, 10, 11, 9, 13, 10, 6, 11, 8, 10, 8, 7, 6, 11, 8, 20)
```

Visualize distribution of data:

```{r}
hist(speed)
```


Find the mean of speed and a 95% confidence interval on the mean.

```{r}
t.test(speed)
```

To get just the confidence interval:

```{r}
tout <- t.test(speed)
tout$conf.int
```

This can also be found using an intercept-only linear model:

```{r}
m <- lm(speed ~ 1)
confint(m)
```

Does the sample appear to be from a Normal distribution?

```{r}
qqnorm(speed)
qqline(speed)
```

Using the model output:

```{r}
plot(m, which = 2)
```

A couple of observations seem unusual: 1, 3, 21

```{r}
speed[c(1, 3, 21)]
```

There was a fast reader and a couple of slower readers.



### References
