---
title: "Linear Modeling"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


### Linear model with no predictors (intercept only)

A new reading program is being evaluated at an elementary school. A random sample of 20 students were tested to determine reading speed. Speed was measured in minutes. [@ott2004]

```{r}
speed <- c(5, 7, 15, 12, 8, 7, 10, 11, 9, 13, 10, 6, 11, 8, 10, 8, 7, 6, 11, 8, 20)
```

Visualize distribution of data:

```{r}
hist(speed)
```


Find the mean of speed and a 95% confidence interval on the mean.

```{r}
t.test(speed)
```

To get just the confidence interval:

```{r}
tout <- t.test(speed)
tout$conf.int
```

This can also be found using an intercept-only linear model:

```{r}
m <- lm(speed ~ 1)
confint(m)
```

Does the sample appear to be from a Normal distribution?

```{r}
qqnorm(speed)
qqline(speed)
```

Using the model output:

```{r}
plot(m, which = 2)
```

A couple of observations seem unusual: 1, 3, 21

```{r}
speed[c(1, 3, 21)]
```

There was a fast reader and a couple of slower readers.


### Linear model with categorical predictors


The `schooldays` data frame from the HSAUR3 package has 154 rows and 5 columns. It contains data from a sociological study of Australian Aboriginal and white children. Model `absent` (number of days absent during school year) as a function of `race`, `gender`, `school` (school type), and `learner` (average or slow learner). [@hsaur3] 

```{r}
library(HSAUR3)
data("schooldays")
```


### Linear model with categorical and numeric predictors

The `bp.obese` data frame from the ISwR package has 102 rows and 3 columns. It contains data from a random sample of Mexican-American adults in a small California town. Analyze blood pressure (`bp`) as a function of obesity (`obese`) and gender (`sex`). Here `obese` is a ratio of actual weight to ideal weight from New York Metropolitan Life Tables. [@ISwR] 

```{r}
library(ISwR)
data("bp.obese")
```




### References
