---
title: "Logistic Regression Models"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Inference about a proportion

100 Hershey's Kisses are put into a large cup and dumped on a table. We want to estimate the proportion of Kisses we can expect to land on their base. [@bilder2015] In our experiment we observe 39 out of 100. What is the 95% confidence interval on the proportion?

```{r}
prop.test(x = 39, n = 100)
```

To get just the confidence interval:

```{r}
pout <- prop.test(x = 39, n = 100)
pout$conf.int
```

This can be analyzed as an intercept-only binary logistic regression model. The confidence interval is slightly different.

```{r}
m <- glm(cbind(39, 100-39) ~ 1, family = binomial)
plogis(confint(m))
```

There are many confidence intervals that can be calculated for proportions. The binom package calculates 11 of them. [@binom]

```{r}
library(binom)
binom.confint(x = 39, n = 100)
```

The "prop.test" and "profile" versions are the two we calculated above. 

### References
